---
title: "Assignment2"
author: "Austin Thebner"
output: html_document
---

library(class)
library(caret)

## Partition the Data

First, import the Universal Bank dataset.

```{r}
# Import the dataset
bank <- read.csv("UniversalBank.csv")

# View the first few rows
head(bank)



# Determine the number of observations
n <- nrow(bank)

# Randomly select 60% of the observations for training
train_index <- sample(1:n, size = 0.60 * n)

# Create the training dataset
training <- bank[train_index, ]

# Create the validation dataset
validation <- bank[-train_index, ]

# Check the number of observations in each dataset
nrow(training)
nrow(validation)
```


```{r}
## k-NN Classification

library(class)



# Partition data into training (60%) and validation (40%)
set.seed(123)

train_index <- sample(
  1:nrow(bank),
  size = 0.60 * nrow(bank)
)

training <- bank[train_index, ]
validation <- bank[-train_index, ]

# --------------------------------------------------
# Create dummy variables for Education
# --------------------------------------------------

bank$Education_1 <- ifelse(bank$Education == 1, 1, 0)
bank$Education_2 <- ifelse(bank$Education == 2, 1, 0)
bank$Education_3 <- ifelse(bank$Education == 3, 1, 0)

# --------------------------------------------------
# Select predictors
# Exclude ID, ZIP Code, and Personal Loan
# --------------------------------------------------

predictors <- bank[, c(
  "Age",
  "Experience",
  "Income",
  "Family",
  "CCAvg",
  "Mortgage",
  "Securities.Account",
  "CD.Account",
  "Online",
  "CreditCard",
  "Education_1",
  "Education_2",
  "Education_3"
)]

# --------------------------------------------------
# Standardize predictors
# --------------------------------------------------

predictors_scaled <- scale(predictors)

# --------------------------------------------------
# Create the customer to classify
# --------------------------------------------------

new_customer <- data.frame(
  Age = 40,
  Experience = 10,
  Income = 84,
  Family = 2,
  CCAvg = 2,
  Mortgage = 0,
  Securities.Account = 0,
  CD.Account = 0,
  Online = 1,
  CreditCard = 1,
  Education_1 = 0,
  Education_2 = 1,
  Education_3 = 0
)

# Make sure variables are in the same order
new_customer <- new_customer[, names(predictors)]

# Standardize the new customer
new_customer_scaled <- scale(
  new_customer,
  center = attr(predictors_scaled, "scaled:center"),
  scale = attr(predictors_scaled, "scaled:scale")
)

# --------------------------------------------------
# k-NN classification with k = 1
# --------------------------------------------------

knn_result <- knn(
  train = predictors_scaled[train_index, ],
  test = new_customer_scaled,
  cl = factor(bank$Personal.Loan[train_index]),
  k = 1,
  prob = TRUE
)

# Display prediction
knn_result

# Display classification
if (knn_result == 1) {
  print("Loan Acceptance: 1")
} else {
  print("Loan Rejection: 0")
}

# Display probability
probability <- attr(knn_result, "prob")
probability

# Display classification
if (knn_result == 1) {
  print("Loan Acceptance: 1")
} else {
  print("Loan Rejection: 0")
}

# Display probability
probability <- attr(knn_result, "prob")
probability
```


```{r}
## Choosing k

# Test different values of k
k_values <- seq(1, 20, by = 1)

# Store validation error rates
error_rates <- numeric(length(k_values))

for (i in seq_along(k_values)) {
  
  # Run k-NN using the training data
  knn_pred <- knn(
    train = predictors_scaled[train_index, ],
    test = predictors_scaled[-train_index, ],
    cl = factor(bank$Personal.Loan[train_index]),
    k = k_values[i]
  )
  
  # Calculate validation error
  error_rates[i] <- mean(
    knn_pred != factor(bank$Personal.Loan[-train_index])
  )
}

# Create a table of k values and error rates
k_results <- data.frame(
  k = k_values,
  Error_Rate = error_rates
)

k_results

# Find the k with the lowest validation error
best_k <- k_values[which.min(error_rates)]

best_k
```


```{r}
## Confusion Matrix for the Validation Data

# Perform k-NN using the best k
validation_pred <- knn(
  train = predictors_scaled[train_index, ],
  test = predictors_scaled[-train_index, ],
  cl = factor(bank$Personal.Loan[train_index]),
  k = best_k
)

# Create the confusion matrix
confusion_matrix <- table(
  Actual = factor(bank$Personal.Loan[-train_index]),
  Predicted = validation_pred
)

# Display the confusion matrix
confusion_matrix
```


```{r}
## Classify New Customer Using the Best k

# Create the new customer
new_customer <- data.frame(
  Age = 40,
  Experience = 10,
  Income = 84,
  Family = 2,
  CCAvg = 2,
  Mortgage = 0,
  Securities.Account = 0,
  CD.Account = 0,
  Online = 1,
  CreditCard = 1,
  Education_1 = 0,
  Education_2 = 1,
  Education_3 = 0
)

# Make sure the variables are in the same order as the predictors
new_customer <- new_customer[, names(predictors)]

# Standardize the new customer's predictors
new_customer_scaled <- scale(
  new_customer,
  center = attr(predictors_scaled, "scaled:center"),
  scale = attr(predictors_scaled, "scaled:scale")
)

# Classify the customer using the best k
customer_prediction <- knn(
  train = predictors_scaled[train_index, ],
  test = new_customer_scaled,
  cl = factor(bank$Personal.Loan[train_index]),
  k = best_k
)

# Display the predicted class
customer_prediction

# Display the classification in words
if (customer_prediction == 1) {
  print("The customer is classified as a loan acceptance (1).")
} else {
  print("The customer is classified as a loan rejection (0).")
}